Hyperbolic SMC: a Variational Proposal for Sequential Monte Carlo Phylogenetic Inference

### 2024-07-31 paper/paper.tex
Fix algo
Write about algorithm

Algorithm \ref{alg:hyperbolic-smc} outlines how the Hyperbolic SMC proposal fits into the VCSMC framework.

% TODO replace this with Hyperbolic SMC
\begin{algorithm}
  \caption{Hyperbolic Sequential Monte Carlo}
  \label{alg:hyperbolic-smc}
  \begin{algorithmic}[1]
    \State \textbf{Input:} $\theta = (Q, v_0^{1:N}, \sigma)$
    \State Initialize $\forall k \in [K], s_0^k \leftarrow \bot, w_0^k \leftarrow 1/K$
    \State Initialize $\forall k \in [K], \forall n \in [N], v_0^{k,n} \leftarrow v_0^n$
    \For{$r = 1$ \textbf{to} $R = N - 1$}
    \For{$k \in [K]$}
    \State Resample particle
    \[
      \mathbb{P}(i_\text{resample} = i) = \frac{w_{r-1}^i}{\sum_{l=1}^K w_{r-1}^l} \; , \forall \, i \in [K]
    \]
    \State Sample two trees uniformly
    \[
      \mathbb{P}(t_\text{left} = i, t_\text{right} = j) = \frac{1}{\binom{N-r+1}{2}} , \; \forall \, i < j
    \]
    \State Form internal embedding $v_\text{parent}$ from $v_{r-1}^{k,t_\text{left}}, v_{r-1}^{k,t_\text{right}}$
    \State Compute parent-child distances
    \[
      d_\text{left} = d\qty(v_{r-1}^{k,t_\text{left}}, v_\text{parent}) \;,\; d_\text{right} = d\qty(v_{r-1}^{k,t_\text{right}}, v_\text{parent})
    \]
    \State Sample branch lengths from log-normal distributions (with reparameterization trick)
    \[
      b_\text{left} = \exp(\log(d_\text{left}) + \sigma Z_\text{left}) \;,\; b_\text{right} = \exp(\log(d_\text{right}) + \sigma Z_\text{right})
    \]
    \[
      Z_\text{left}, Z_\text{right} \sim \mathcal{N}(0, 1)
    \]
    \State Form new partial state $s_r^k$ by merging trees $t_\text{left}, t_\text{right}$
    \State Compute proposal probability
    \[
      q(s_r^k | s_{r-1}^{i_\text{resample}}) = \mathbb{P}(t_\text{left} = i, t_\text{right} = j) \cdot f_\text{log-normal}(b_\text{left} | d_\text{left}, \sigma) \cdot f_\text{log-normal}(b_\text{right} | d_\text{right}, \sigma)
    \]
    \State Compute weight
    \[
      w_r^k = \frac{\pi(s_r^k) \cdot \nu^-(s_{r-1}^{i_\text{resample}})}{\pi(s_{r-1}^{i_\text{resample}}) \cdot  q(s_r^k | s_{r-1}^{i_\text{resample}})}
    \]
    \State
    Update tree root embeddings
    \[
      \qty{v_r^{k,1}, \cdots, v_r^{k,N-r}} \leftarrow \qty(\qty{v_{r-1}^{k,1}, \cdots, v_{r-1}^{k,N-r+1}} \setminus \qty{v_{r-1}^{t_\text{left}}, v_{r-1}^{t_\text{right}}}) \cup \qty{v_\text{parent}}
    \]
    \EndFor
    \EndFor
    \State \Return $s_{1:R}^{1:K}, w_{1:R}^{1:K}$
  \end{algorithmic}
\end{algorithm}

>>>>

% TODO cite https://www.stats.ox.ac.uk/~doucet/wang_bouchardcote_doucet_BayesianphylogeneticscombinatorialSMC_JASA2015.pdf
Algorithm \ref{alg:hyperbolic-smc} outlines how the Hyperbolic SMC proposal fits into the VCSMC framework. At each optimization step, the algorithm takes as input $Q$, the (possibly learnable) transition matrix, $v_0^{1:N}$, the learned embeddings of the $N$ tree tips, and $\sigma$, the global branch length variance. Following the VCSMC framework, at each step $r$, we resample the $k$ particles, then extend each partial state $s_{r-1}^{i_\text{resample}}$ to $s_r^k$ via the proposal distribution $q(\cdot | s_{r-1}^{i_\text{resample}})$. The two chosen trees dictate the new internal embedding $v_\text{parent}$, which sets the branch length distributions. $\nu^-(\cdot)$ corrects for the overcounting problem detailed in [Wang et al]. As in the original VCSMC framework, an unbiased estimate for the marginal likelihood is
\begin{align*}
  \hat{\mathcal{Z}}_\text{CSMC} = \prod_{r=1}^R \qty(\frac{1}{K} \sum_{k=1}^K w_r^k) \; .
\end{align*}

% TODO replace this with Hyperbolic SMC
\begin{algorithm}
  \caption{Hyperbolic Sequential Monte Carlo}
  \label{alg:hyperbolic-smc}
  \begin{algorithmic}[1]
    \State \textbf{Input:} $\theta = (Q, v_0^{1:N}, \sigma)$
    \State Initialize $\forall k \in [K], s_0^k \leftarrow \bot, w_0^k \leftarrow 1/K$
    \State Initialize $\forall k \in [K], \forall n \in [N], v_0^{k,n} \leftarrow v_0^n$
    \For{$r = 1$ \textbf{to} $R = N - 1$}
    \For{$k \in [K]$}
    \State Resample particle
    \[
      \mathbb{P}(i_\text{resample} = i) = \frac{w_{r-1}^i}{\sum_{l=1}^K w_{r-1}^l} \; , \forall \, i \in [K]
    \]
    \State Sample two trees uniformly
    \[
      \mathbb{P}(t_\text{left} = i, t_\text{right} = j) = \frac{1}{\binom{N-r+1}{2}} , \; \forall \, i < j
    \]
    \State Form internal embedding $v_\text{parent}$ from $v_{r-1}^{k,t_\text{left}}, v_{r-1}^{k,t_\text{right}}$
    \State Compute parent-child distances
    \[
      d_\text{left} = d\qty(v_{r-1}^{k,t_\text{left}}, v_\text{parent}) \;,\; d_\text{right} = d\qty(v_{r-1}^{k,t_\text{right}}, v_\text{parent})
    \]
    \State Sample branch lengths from log-normal distributions (with reparameterization trick)
    \[
      b_\text{left} = \exp(\log(d_\text{left}) + \sigma Z_\text{left}) \;,\; b_\text{right} = \exp(\log(d_\text{right}) + \sigma Z_\text{right})
    \]
    \[
      Z_\text{left}, Z_\text{right} \sim \mathcal{N}(0, 1)
    \]
    \State Form new partial state $s_r^k$ by merging trees $t_\text{left}, t_\text{right}$
    \State Compute proposal probability
    \[
      q(s_r^k | s_{r-1}^{i_\text{resample}}) = \mathbb{P}(t_\text{left} = i, t_\text{right} = j) \cdot f_\text{log-normal}(b_\text{left} | d_\text{left}, \sigma) \cdot f_\text{log-normal}(b_\text{right} | d_\text{right}, \sigma)
    \]
    \State Compute weight
    \[
      w_r^k = \frac{\pi(s_r^k) \cdot \nu^-(s_{r-1}^{i_\text{resample}})}{\pi(s_{r-1}^{i_\text{resample}}) \cdot  q(s_r^k | s_{r-1}^{i_\text{resample}})}
    \]
    \State
    Update tree root embeddings
    \[
      \qty{v_r^{k,1}, \cdots, v_r^{k,N-r}} \leftarrow \qty(\qty{v_{r-1}^{k,1}, \cdots, v_{r-1}^{k,N-r+1}} \setminus \qty{v_{r-1}^{k,t_\text{left}}, v_{r-1}^{k,t_\text{right}}}) \cup \qty{v_\text{parent}}
    \]
    \EndFor
    \EndFor
    \State \Return $s_{1:R}^{1:K}, w_{1:R}^{1:K}$
  \end{algorithmic}
\end{algorithm}
